The ``Dark Side'' of Phylogenetic Comparative Methods$^\dag$

### 2015-04-14 Manuscript/dark.tex
Minor edits

\title{The ``Dark Side'' of Phylogenetic Comparative Methods$^\dag$}
\author{
  Natalie Cooper$^{1,2,*}$, Matthew W. Pennell$^{3}$ \\
  and Richard G. FitzJohn$^{4}$ 
  %Cecile An\'{e}$^{2}$, Wayne Maddison$^{3}$,\\  Gavin H. Thomas$^{5}$, April Wright$^{6}$,\\
}
\date{}
\affiliation{\noindent{\footnotesize

 % $^2$ Wisconsin\\
 % $^3$ UBC\\
 % $^5$ Sheffield\\
 % $^6$ Austin\\
  $^1$ School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland.\\ 
  $^2$ Trinity Centre for Biodiversity Research, Trinity College Dublin, Dublin 2, Ireland.\\
  $^3$ Institute for Bioinformatics and Evolutionary Studies, University of Idaho, Moscow, ID 83844, U.S.A.\\
  $^4$ Department of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia. \\
  $^*$ Corresponding author: ncooper@tcd.ie; Zoology Building, Trinity College Dublin, Dublin 2, Ireland. 
       Fax: +353 1 677 8094; Tel: +353 1 896 1926.\\
  $^\dag$This paper is based on a Society of Systematic Biologists symposium at Evolution 2014, in Raleigh NC.\\
}}

@@

One major issue is that two groups of people are involved with PCMs. 
At one extreme, methods-developers (``wookies'') develop new methods and implement them. They are (generally) aware of the limitations of their methods, and the assumptions that underly them. 
At the other extreme, na\"{i}ve methods-users (``ewoks'') just want to use PCMs to answer empirical questions with their data and, although they have expertise in other biological domains, they may not be aware of issues with the methods (of course the majority of people fall somewhere between these two extremes on the ``wookie-ewok'' spectrum; many methods-developers are developing methods with a specific empirical question in mind, while there are many advanced methods-users who have detailed knowledge of the limitations of the methods they use). 
The problem is that although methods-developers (and advanced methods-users) are aware of the problems with PCMs, this information is not being effectively transferred to methods-users. 
Additionally, the tools and approaches used to fit models are often far more user-friendly and better documented than the methods used to to assess whether that model fit is reasonable. 
Clearly more effort is needed to bridge the widening gap between methods-users and methods-developers. Here we explore the causes of this communication gap and suggest some potential solutions. % NC: still not happy with this final line or two.

@@

%Note that we assume that methods-users ignore assumptions and biases because they are unaware of them. Of course, there may be methods-users who are perfectly aware of these issues but choose to sweep them under the carpet to get their work published. Although we believe there is no excuse for doing this, we also recognize that the ``publish or perish'' culture of academia may pressurize nontenured scientists into this. Hopefully some of the changes we suggest below will also help with this problem. 
% NC: This could be left out!

@@

Hopefully by trying to implement some of these solutions we can banish the ``Dark Side'' of PCMs and create a ``New Hope'' for the whole PCM community, whether they be wookies, ewoks or somewhere in between.

>>>>

\title{The ``Dark Side'' of Phylogenetic Comparative Methods$^\dag$}
\author{
  Natalie Cooper$^{1,2,*}$, Gavin H. Thomas$^{3}$ \\
  and Richard G. FitzJohn$^{4}$ 
  %Cecile An\'{e}$^{2}$, Wayne Maddison$^{3}$,\\  $^{5}$, April Wright$^{6}$,\\
}
\date{}
\affiliation{\noindent{\footnotesize

  $^1$ School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland.\\ 
  $^2$ Trinity Centre for Biodiversity Research, Trinity College Dublin, Dublin 2, Ireland.\\
  $^3$ Department of Animal and Plant Sciences, University of Sheffield, Sheffield S10 2TN, UK.\\
  $^4$ Department of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia. \\
  $^*$ Corresponding author: ncooper@tcd.ie; Zoology Building, Trinity College Dublin, Dublin 2, Ireland. 
       Fax: +353 1 677 8094; Tel: +353 1 896 1926.\\
  $^\dag$This paper is based on a Society of Systematic Biologists symposium at Evolution 2014, in Raleigh NC.\\
}}

@@

One major issue is that two groups of people are involved with PCMs. 
At one extreme, methods-developers develop new methods and implement them. They are (generally) aware of the limitations of their methods, and the assumptions that underly them. 
At the other extreme, na\"{i}ve methods-users just want to use PCMs to answer empirical questions with their data and, although they have expertise in other biological domains, they may not be aware of issues with the methods (of course the majority of people fall somewhere between these two extremes on this spectrum; many methods-developers are developing methods with a specific empirical question in mind, while there are many advanced methods-users who have detailed knowledge of the limitations of the methods they use). 
The problem is that although methods-developers (and advanced methods-users) are aware of the problems with PCMs, this information is not being effectively transferred to methods-users. 
Additionally, the tools and approaches used to fit models are often far more user-friendly and better documented than the methods used to to assess whether that model fit is reasonable. 
Clearly more effort is needed to bridge the widening gap between methods-users and methods-developers. Here we explore the causes of this communication gap and suggest some potential solutions. % NC: still not happy with this final line or two.

@@

@@
